Distinct snow algae communities along an elevational gradient

### 2019-10-30 manuscript/Casey1.Rmd
re outlined intro, still mid-way thru

Snow algae are cold-adapted microalgae that bloom in melting summer snow [@Leya2013]. Red pigments produced by some species of snow algae can color vast areas of polar and alpine snowfields, which decreases snow surface albedo---by one estimate, enough to substantially increase the rate of snow melt [@Ganey2017]. Earlier snow melt exposes darker surfaces below, contributing to snow-albedo feedback and global warming [@Dery2007], and loss of snowpack reservoirs can threaten water supplies, leading to drought [@Staudinger2014;@Barnhart2016]. Thus, snow algae could play an important role in snow melt regulation. Snow algae are the primary producers in a microbiome that includes grazers, fungi, and bacteria [@Hoham2001], but whether this microbiome can be considered as a single microbiome, or whether there are distinctions between types of snow algae blooms has been poorly studied. The first step in understanding this microbiome is defining the algae species that are present in snow algae blooms, and their community structure.

Until recently, snow algae species were defined and identified on the basis of morphology. Morphology can be deceptive when identifying microalgae, as the same species can look completely different from one life stage to the next [@Shah2016;@Salome2019]. In the lab, cultures of *Chloromonas* are green biflagellates, but in the field the same species contain yellow or orange pigment, and thick cell walls with flanges or spikes. Conversely, different genetic species can look nearly identical: for example, cell types formerly known as *Chloromonas cf. nivalis* are now split into multiple species [@Matsuzaki2019;@Prochazhova2019a].

To date, genetic studies have identified three main genera of algae that cause snow algae blooms: *Sanguina*, *Chlainomonas*, and *Chloromonas*. Neither of the first two have cultured representatives, but both produce the red pigment astaxanthin [@Remias2005;Remias2016]. *Sanguina* was only recently designated and contains just two species, one of which *Sanguina nivaloides* is found in alpine and polar snowfields worldwide, and is likely synonymous with some of *Chlamydomonas cf. nivalis* [@Prochazkova2018c]. *Chlainomonas* also produce red pigment, but are physically much larger than *Sanguina* (up to 40 $\mu$m diameter). Although *Chlainomonas* look very different from *Chloromonas*, *rbc*L and 18S phylogeny places them within the *Chloromonas* clade [@Novis2008]. *Chloromonas* are highly diverse, with at least 13 named species [@Matsuzaki2019]; in contrast, only two species are currently described in *Chlainomonas* [@Novis2008].

Snow algae grow in the liquid water surrounding rounded snow granules [@Leya2013]. In this micro-habitat, temperature is nearly constant, but snow moisture, pH, nutrients, and light can differ greatly within the snowpack depending on location of snow and time of year [@Loranger1988;@Kuhn2001;@Gorton2007;@Ganey2017]. Differences in snow algae community composition could offer insight into the environmental factors that determine snow algae distribution. Green snow blooms dominated by *Chloromonas* are reported from forested sites and shaded snowbanks [@Terashima2017;@Remias2013a], although some *Chloromonas* may be specially adapted for high light intensity in the alpine [@Prochazkova2019a]. Two studies using microscopy found that red snow containing *Chlamydomonas cf. nivalis* was prevalent in the above timberline, while green and orange snows caused by *Chloromonas cf. brevispina* and *Chloromonas cf. nivalis* were prevalent in conifer forests below timberline [@Hoham1979a;@Nedbalova2008a]. Differences in snow moisture could influence which species of snow algae grow; several species of snow algae are anecdotally reported to only inhabit very saturated, waterlogged snow [@Lutz2015;@Remias2013;@Novis2008;@Prochazkova2018c], while  *Chlamydomonas cf. nivalis* is hypothesized to be specially adapted to well-drained, drier snow [@Ganey2018].

Microscopic observation shows that snow algae blooms can contain a diversity of cell morphologies in the same sample [@Lutz2018], but the genetic community structure of snow algae may not correlate with what is seen under the microscope. 18S amplicon high-throughput sequencing (HTS) shows that snow algae may contain more diversity than meets the eye: although red snow from sites across the Arctic were dominated by uncultured Chlamydomonadaceae, they also contained lower abundance of reads assigned to *Chloromonas* and *Raphidonema* [@Lutz2016a]. 18S is highly conserved and therefore suited to survey broadly across a variety of taxa; however, due to its high similarity between closely related species short segments of thie gene are poorly-differentiated among closely related taxa (ref). The *rbc*L plastid gene coding for the large subunit of RuBisCO is highly differentiated among photosynthetic organisms, and contains reference sequences for snow algae, making it an ideal complement to survey snow algae communities.

In this study, we surveyed snow algae communities from the mountains surrounding Vancouver, BC, Canada using light microscopy, 18S and *rbc*L amplicon HTS. Our aims were to: 1) see what species were present, 2) compare community composition between samples. We found ...

@@

**Snow algae community composition co-varies with date and elevation**. *Sanguina* was more abundant in alpine sites, as well as one OTU of *Chlainomonas* and two OTUs of *Chloromonas*, while *Chloromonas krienitzii* and Chlainomonas OTU6 were found in high relative abundance below treeline (Fig. 2, 3). This consistent with the microscopy observations of @Hoham1979a and @Nedbalova2008a, who found primarily *Chlamydomonas cf. nivalis* above treeline and *Chloromonas* below treeline. However, in microscopy surveys we infrequently observed *Sanguina* below treeline, and *Chloromonas krienitzii* above treeline. We found higher diversity in samples from higher elevation, both at the ASV level and OTU level<!--Chao1 w fossil.-->, with several OTUs that were not found at lower elevations. *Raphidonema* and related ASVs were restricted to high alpine sites, with the highest relative abundance occurring in samples from August and September. This fits with the interpretation of @Stibal2005, who suggest that *Raphidonema* is an opportunistic soil algae that blows into snow fields from surrounding dirt to colonize the snow. However, rare taxa should not be discounted as playing an important role in ecosystem functioning [@Jousset2017].

>>>>

Snow algae are cold-adapted microalgae that bloom in melting summer snow throughout alpine and polar regions [@Leya2013]. Red snow algae blooms can cover vast areas of snowfield, which decreases snow surface albedo---by one estimate, enough to substantially increase the rate of snow melt [@Ganey2017]. Earlier snow melt exposes darker surfaces below, contributing to snow-albedo feedback and global warming [@Dery2007], which can threaten water supplies [@Staudinger2014] and worsten drought [@Barnhart2016]. Thus, snow algae could play an important role in snow melt regulation, but little is known about how these blooms form. We don't know the spatial extent of snow algae blooms worldwide, what conditions trigger snow algae bloom formation, or how cells arrive on the surface of the snow to "seed" the bloom. Critically, we lack basic understanding of the species composition and diversity in snow algae blooms, which is fundamental to answering these long-standing mysteries about snow algae blooms.

Historically snow algae were classified by cell morphology, however, morphology can be deceptive when identifying microalgae. Until recently, all red cells from red snow were referred to as *Chlamydomonas cf. nivalis* [@Remias2013;@Segawa2018], desipte the fact that this morphology-based species was thought to contain multiple distinct species [@Kol1961]. Recent genetic work designated found that some of these cells were not closely related to *Chlamydomonas* at all, and were designated as a new genus and species: *Sanguina nivaloides* [@Prochazkova2018c]. Other genetic work shows that morphology-based species of snow algae may in fact contain multiple genetically unrelated clades: for example, *Chloromonas cf. nivalis* contains at least four distinct clades [@Matsuzaki2019].

<!-- hammer this home: WE LACK BASIC UNDERSTANDING ABOUT SPECIES COMPOSITOIN AND DIVERSITY -->

<!-- sequence cultures, and batch, and single cell, but this dosent tell us whats actually in the field -->
More recently, snow algae taxonomy has

<!-- high throughput: lacking ref data, no otu based, none in alpine environments (despite this tantalizing evidence of microscopy based, etc) -->

<!-- But no one until now has used high throughput sequencing to compare snow algae blooms in alpine environments, or used higher resolution marker genes to do so -->

<!-- so we did this... -->
In this study, we surveyed snow algae communities from the mountains surrounding Vancouver, BC, Canada using light microscopy, 18S and *rbc*L amplicon HTS. Our aims were to: 1) see what species were present, 2) compare community composition between samples. We found ...

<!-- the red green problem -->
The same species can look completely different depending on the life stage [@Shah2016;@Salome2019]; conversely, different species can look nearly identical. For example, cell morphologies known as *Chloromonas cf. nivalis* appear in at least four different lineages. One of these lineages, *Chloromonas muramotoi* appear as green biflagellates in culture, but in the field the same species features thick cell walls with flanges [@Matsuzaki2019]. However, life cycle changes in the the cosmopolitan and widespread genus *Sanguina* have yet to be documented: despite over a century of study [@Kol1961] no one has sucessfully cultured this genus.

<!-- habitat differences -->
Snow algae grow in the liquid water surrounding rounded snow granules [@Leya2013]. In this micro-habitat, temperature is nearly constant, but snow moisture, pH, nutrients, and light can differ greatly within the snowpack depending on location of snow and time of year [@Loranger1988;@Kuhn2001;@Gorton2007;@Ganey2017]. Differences in snow algae community composition could offer insight into the environmental factors that determine snow algae distribution. Green snow blooms dominated by *Chloromonas* are reported from forested sites and shaded snowbanks [@Terashima2017;@Remias2013a], although some *Chloromonas* may be specially adapted for high light intensity in the alpine [@Prochazkova2019a]. Two studies using microscopy found that red snow containing *Chlamydomonas cf. nivalis* was prevalent in the above timberline, while green and orange snows caused by *Chloromonas cf. brevispina* and *Chloromonas cf. nivalis* were prevalent in conifer forests below timberline [@Hoham1979a;@Nedbalova2008a]. Differences in snow moisture could influence which species of snow algae grow; several species of snow algae are anecdotally reported to only inhabit very saturated, waterlogged snow [@Lutz2015;@Remias2013;@Novis2008;@Prochazkova2018c], while  *Chlamydomonas cf. nivalis* is hypothesized to be specially adapted to well-drained, drier snow [@Ganey2018].

<!-- habitat differences, short new form -->
  Red snow is also occasionally reported to be caused by another red-pigmented algae: *Chlainomonas* [@Novis2008]. Little is known about the distribution of this genus; one study suggested that *Chlainomonas* is limited to high-moisture snowpack above alpine lakes [@Prochazkova2018]. Green snow may contain a variety of algae, notably *Chloromonas* (refs), as well as algae from other families including Trebouxiophyceae and Chrysophyceae (refs). Morphology-based cell counts suggest that green snow caused by *Chloromonas* is prevalent in forested areas and red snow containing *Chlamydomonas cf. nivalis* is prevalent in alpine areas [@Hoham1979a;@Nedbalova2008a]; however, morphology can be misleading when identifying microalgae.

Although much of the snow algae literature focusses on eye-catching red snow, different species of snow algae can also color snow green [@Lutz2015].

<!-- hts -->
Microscopic observation shows that snow algae blooms can contain a diversity of cell morphologies in the same sample [@Lutz2018], but the genetic community structure of snow algae may not correlate with what is seen under the microscope. 18S amplicon high-throughput sequencing (HTS) shows that snow algae may contain more diversity than meets the eye: although red snow from sites across the Arctic were dominated by uncultured Chlamydomonadaceae, they also contained lower abundance of reads assigned to *Chloromonas* and *Raphidonema* [@Lutz2016a]. 18S is highly conserved and therefore suited to survey broadly across a variety of taxa; however, due to its high similarity between closely related species short segments of thie gene are poorly-differentiated among closely related taxa (ref). The *rbc*L plastid gene coding for the large subunit of RuBisCO is highly differentiated among photosynthetic organisms, and contains reference sequences for snow algae, making it an ideal complement to survey snow algae communities.

Until recently, snow algae sequences were lacking in reference databases,

@@

**Snow algae community composition co-varies with date and elevation**. *Sanguina* was more abundant in alpine sites, as well as one OTU of *Chlainomonas* and two OTUs of *Chloromonas*, while *Chloromonas krienitzii* and Chlainomonas OTU6 were found in high relative abundance below treeline (Fig. 2, 3). This consistent with the microscopy observations of @Hoham1979a and @Nedbalova2008a, who found primarily *Chlamydomonas cf. nivalis* above treeline and *Chloromonas* below treeline. However, in microscopy surveys we infrequently observed *Sanguina* below treeline, and *Chloromonas krienitzii* above treeline. We found higher diversity in samples from higher elevation, both at the ASV level and OTU level <!--Chao1 w fossil.-->, with several OTUs that were not found at lower elevations. *Raphidonema* and related ASVs were restricted to high alpine sites, with the highest relative abundance occurring in samples from August and September. This fits with the interpretation of @Stibal2005, who suggest that *Raphidonema* is an opportunistic soil algae that blows into snow fields from surrounding dirt to colonize the snow. However, rare taxa should not be discounted as playing an important role in ecosystem functioning [@Jousset2017].
